---
title: "CVB3_TT quality control and clustering"
author: "Christine Mariskanish"
output:
  html_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r libraries}
library(dplyr)
library(patchwork)
library(ggplot2)
library(Seurat)
library(SeuratObject)

## This analysis targets Seurat v5. Install the environment once, outside the
## notebook -- the pre-restructure version carried install_version() calls in a
## chunk, which reinstalled Seurat on every knit:
##
##   conda env create -f workflow/envs/scissors.yaml && conda activate scissors
##
## Originally written against Seurat 4.4.0, now archived on CRAN. This notebook
## builds a single-sample object, so it has one layer and needs no JoinLayers;
## see analysis/PV_mutants_integrated.R for the merged case.
stopifnot(packageVersion("Seurat") >= "5.0.0")

source(file.path(rprojroot::find_root(rprojroot::has_file("config.sh")), "analysis", "config.R"))
source(file.path(rprojroot::find_root(rprojroot::has_file("config.sh")), "analysis", "helpers.R"))

SAMPLE_ID <- "CVB3_TT"
```

Paths come from `analysis/config.R`, which resolves the lab share wherever it
is mounted. The pre-restructure notebooks hardcoded
`/Volumes/LVD_QVE/Projects/PTD_StrandSpecificCounting_scRNAseq/CellRanger`, and mixed that
with `/Volumes/LVD_QVE` in the same file -- a case difference that breaks on
any case-sensitive volume.

# Load counts

```{r load}
matrix_dir <- file.path(CELLRANGER_DIR, SAMPLE_ID, "outs", "filtered_feature_bc_matrix")
stopifnot(dir.exists(matrix_dir))

counts <- Read10X(data.dir = matrix_dir)
obj <- CreateSeuratObject(counts = counts, project = SAMPLE_ID,
                          min.cells = 3, min.features = 10)
rm(counts); gc(verbose = FALSE)
obj
```

## Metadata

`meta.data` carries one row per cell: the sample ID (`orig.ident`), UMIs per
cell (`nCount_RNA`) and genes detected per cell (`nFeature_RNA`). It is
reachable with either `@meta.data` or `[[ ]]`.

```{r metadata}
meta <- obj[[]]
dim(meta)
head(meta)
summary(meta$nCount_RNA)
summary(meta$nFeature_RNA)
```

# QC metrics

```{r qc-metrics}
obj[["percent.mt"]]      <- PercentageFeatureSet(obj, pattern = "^MT-")
obj[["percent.ribo"]]    <- PercentageFeatureSet(obj, pattern = "^RP[SL]")
obj[["percent.mtribo"]]  <- PercentageFeatureSet(obj, pattern = "^MT-|^RP[SL]")
```

Doublet scores come from `scripts/04_scrublet.py`. They are joined **by
barcode** -- the pre-restructure notebooks passed an unnamed vector to
`AddMetaData`, which assigns by position, so any barcode dropped by
`min.features = 10` shifted every subsequent score onto the wrong cell and the
`doublet_scores` filter below then discarded the wrong cells. See
`read_doublet_scores()` in `analysis/helpers.R`.

```{r doublets}
obj <- read_doublet_scores(obj, path = doublet_score_path(SAMPLE_ID),
                           sample_id = SAMPLE_ID)
head(obj[[]])
table(obj$predicted_doublets)
```

## Distributions

```{r qc-plots, fig.width=9, fig.height=4}
ggplot(obj[[]], aes(x = nCount_RNA)) +
  geom_histogram(bins = 40) + theme_bw() +
  labs(x = "Number of UMIs per cell", y = "Number of cells") +
  scale_x_continuous(trans = "log2")

ggplot(obj[[]], aes(x = nFeature_RNA)) +
  geom_histogram(bins = 40) + theme_bw() +
  labs(x = "Number of genes per cell", y = "Number of cells") +
  scale_x_continuous(trans = "log2")

ggplot(obj[[]], aes(x = nFeature_RNA, y = nCount_RNA)) +
  geom_point(pch = ".") + theme_bw() +
  labs(x = "Number of genes per cell", y = "Number of UMIs per cell") +
  scale_x_continuous(trans = "log2") +
  scale_y_continuous(trans = "log2")

ggplot(obj[[]], aes(x = nCount_RNA, y = percent.mtribo)) +
  geom_point(pch = ".") + theme_bw() +
  labs(x = "Number of UMIs per cell",
       y = "Percent mitochondrial or ribosomal") +
  scale_x_continuous(trans = "log2")
```

```{r qc-violin, fig.width=9, fig.height=4}
## ncol matches the number of features plotted; the pre-restructure version
## passed ncol = 5 for three features.
VlnPlot(obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.mtribo"),
        ncol = 4, pt.size = 0.1)
```

# Filtering

Thresholds below were hand-tuned against this sample's distributions above.
They are sample-specific and carried over verbatim from the original notebook.

```{r filter}
obj <- subset(obj, subset = nCount_RNA > 7000 & nFeature_RNA > 2000 & percent.mt < 20)
obj <- subset(obj, subset = doublet_scores < 0.53)
obj

table(obj$predicted_doublets)
FeatureScatter(obj, feature1 = "nCount_RNA", feature2 = "percent.mt")
FeatureScatter(obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
VlnPlot(obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3, pt.size = 0.1)
```

# Normalize and find variable features

```{r normalize}
obj <- NormalizeData(obj)
obj <- FindVariableFeatures(obj, selection.method = "vst", nfeatures = 2000)

top10 <- head(VariableFeatures(obj), 10)
plot1 <- VariableFeaturePlot(obj)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

```{r scale}
all.genes <- rownames(obj)
obj <- ScaleData(obj, features = all.genes)
```

# Cell cycle scoring

Marker sets come from Seurat's built-in `cc.genes` (Tirosh et al. 2015), split
into S and G2/M phase.

The pre-restructure notebook also read
`nestorawa_forcellcycle_expressionMatrix.txt` from `~/Downloads/` into
`exp.mat` and never used it. That read has been dropped -- it was the only
thing tying the analysis to a file in someone's Downloads folder.

```{r cell-cycle}
s.genes   <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

obj <- CellCycleScoring(obj, s.features = s.genes, g2m.features = g2m.genes,
                        set.ident = TRUE)
head(obj[[]])
RidgePlot(obj, features = c("PCNA", "TOP2A", "MCM6", "MKI67"), ncol = 2)
```

Before regression, cells separate almost entirely by cell cycle phase:

```{r cell-cycle-pca}
obj <- RunPCA(obj, features = c(s.genes, g2m.genes))
PCAPlot(obj)

Idents(obj) <- obj$orig.ident
head(Idents(obj))
```

```{r regress-cell-cycle}
obj <- ScaleData(obj, features = all.genes, verbose = TRUE,
                 vars.to.regress = c("S.Score", "G2M.Score"))
```

# Dimensionality reduction and clustering

```{r pca}
obj <- RunPCA(obj, verbose = FALSE)
ElbowPlot(obj)
```

```{r cluster}
obj <- FindNeighbors(obj, dims = 1:20, verbose = FALSE)
obj <- FindClusters(obj, resolution = seq(0.1, 1, by = 0.05), verbose = FALSE)

if (requireNamespace("clustree", quietly = TRUE)) {
  print(clustree::clustree(obj))
} else {
  message("clustree is not installed; skipping the cluster-stability plot.")
}
head(obj[[]])
```

```{r umap, fig.width=9, fig.height=4}
obj <- RunUMAP(obj, dims = 1:20, verbose = FALSE)

for (res in c(0.1, 0.15, 0.2, 0.25, 0.3, 0.35, 0.4, 0.45, 0.5)) {
  print(DimPlot(obj, group.by = sprintf("RNA_snn_res.%s", res)) +
          labs(title = sprintf("resolution %s", res)))
}

RESOLUTION <- 0.25
resolution_column <- sprintf("RNA_snn_res.%s", RESOLUTION)
stopifnot(resolution_column %in% colnames(obj[[]]))
Idents(obj) <- obj[[resolution_column]][, 1]

DimPlot(obj, label = TRUE)
```

# Outputs

```{r write-outputs}
dir.create(RESULTS_DIR, recursive = TRUE, showWarnings = FALSE)

write.csv(as.data.frame(Embeddings(obj, reduction = "umap")),
          file = file.path(CELLRANGER_DIR, SAMPLE_ID, "UMAP.csv"))
write.csv(obj[[]], file = file.path(RESULTS_DIR, paste0(SAMPLE_ID, ".csv")))
saveRDS(obj, file = file.path(RESULTS_DIR, paste0(SAMPLE_ID, ".rds")))
```

# Cluster markers

```{r markers}
markers <- FindAllMarkers(obj, only.pos = TRUE)
write.csv(markers,
          file = file.path(RESULTS_DIR, paste0(SAMPLE_ID, ".markers_Pos.csv")))

## FindAllMarkers returns a zero-row frame -- with no columns at all, so not
## even `cluster` -- when the chosen resolution yields a single cluster or no
## gene reaches significance. Grouping on `cluster` then fails with
## "Column `cluster` is not found", which points at dplyr rather than at the
## clustering. Check first and say what actually happened.
if (nrow(markers) == 0L || !"cluster" %in% colnames(markers)) {
  warning("FindAllMarkers found no markers at resolution ", RESOLUTION,
          " (", nlevels(Idents(obj)), " cluster(s)); skipping the heatmap.\n",
          "  Raise RESOLUTION, or check that this sample has real structure.",
          call. = FALSE)
} else {
  markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10_markers

  print(DoHeatmap(obj, features = top10_markers$gene) + NoLegend())
}
```

```{r marker-panel, fig.width=10, fig.height=8}
FeaturePlot(obj, features = MARKER_PANEL)
```
